Open Data, Open Source and Open Standards in chemistry - The Blue Obelisk five years on

### 2011-04-29 paper.tex
Initial comments.

\author{Charles A Darwin\correspondingauthor$^{1,2}$%
       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
      \and 
         Egon L Willighagen$^3$%
         \email{Egon L Willighagen - egon.willighagen@ki.se}
     \and 
         Rajarshi Guha$^4$%
         \email{Rajarshi Guha - guhar@mail.nih.gov}
	 \and 
	     Christoph Steinbeck$^6$%
	         \email{Christoph Steinbeck - steinbeck@ebi.ac.uk}
     \and 
         Noel M O'Boyle$^5$%
         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
     \and 
         Ola Spjuth$^9$%
         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
     \and 
         Jonathan Alvarsson$^9$%
         \email{Jonathan Alvarsson - jonathan.alvarsson@farmbio.uu.se}
     \and 
         Peter Murray-Rust$^7$%
         \email{Peter Murray-Rust - pm286@cam.ac.uk}
     \and 
         Daniel M Lowe$^7$%
         \email{Daniel Lowe - dl387@cam.ac.uk}
     \and
         Robert Hanson$^8$%
         \email{Robert Hanson - hansonr@stolaf.edu}
      \and
         Jane E Doe\correspondingauthor$^2$%
         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
      \and
         Dmitry Pavlov$^8$%
         \email{Dmitry Pavlov - dpavlov@ggasoftware.com}
      }

@@

\address{%
    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
        Waterloo Road, London, UK\\
    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden\\
    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland\\
    \iid(6)Cheminformatics and Metabolism Team, European Bioinformatics Institute (EBI), Wellcome Trust Genome Campus, Hinxton, Cambridge, UK\\
    \iid(7)Unilever Centre for Molecular Science Informatics, Department of Chemistry, University of Cambridge, Lensfield Road, CB2 1EW, UK\\
    \iid(8)GGA Software Services LLC, 41 Nab. Chernoi rechki 194342, Saint Petersburg, Russia\\
    \iid(9)Department of Pharmaceutical Biosciences, Uppsala University, Box 591, 751 24 Uppsala, Sweden 
}%

@@

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
        \paragraph*{Background:} The Blue Obelisk movement was established in 2005 as a 
		response to the lack of open data, open standards and
		open source (ODOSOS) in chemistry. While other scientific disciplines
		such as physics, biology and astronomy (to name a few) were embracing
		new ways of doing science and reaping the benefits of community
		efforts, there was little if any innovation in the field of chemistry
		and scientific progress was actively hampered by the lack of access to
		data and tools.

        \paragraph*{Results:} This
		contribution looks back on the past 5 years and surveys progress and
		remaining challenges in the areas of Open Source, Open Data and Open
		Standards in chemistry.

        \paragraph*{Conclusions:} Here we show that the Blue Obelisk has been very successful
		in bringing together researchers and developers with common interests
		in ODOSOS, leading to development of many useful resources freely
		available to the chemistry community. But how best to engage with the
		wider chemistry community outside of the Blue Obelisk remains an open
		question.
\end{abstract}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk movement was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.
Since 2005 it has become evident that a good amount of development in open 
chemical information and information is driven by the demands of neighbouring
scientific fields. In many areas in biology, for example, the importance of 
small molecules and their interactions and reactions in biological systems
has been realised. In fact, one of the first free and open databases and ontologies 
of small molecules was created as a resource about chemical structure and nomenclature
by biologists \cite{DeMatos:2009p3839}.
Open workflow systems such as Taverna \cite{Hull:2006p60} and KNIME\cite{WebKNIME}
have opened the door for 
an easy and seamless integration of BO components into executable workflows 
or data pipelines. A combination of the Chemistry Development Kit and Taverna, for instance, was
reported in 2010 \cite{Kuhn:2010p4001}.

KNIME currently has a built-in collection of CDK-based and
OpenBabel-based nodes\cite{WebKNIMEExtensions} for implementing
chemistry workflows. Another collection of chemistry nodes---based on
the RDKit and its Java wrapper---appeared in 2010 and is available to
KNIME users from the KNIME's official ``Community Updates'' site.
Indigo nodes for KNIME are in preparation and will appear nearby the
RDKit nodes by May 2011.

The formation of the Blue Obelisk group is somewhat unusual in that it
is not a funded network, nor does it follow the industry consortium
model. Rather it is a grassroots organisation, catalysed by an initial
core of interested scientists, but with membership open to all who
share one or more of the goals of the group:
\begin{itemize}
\item Open Source. One can use other people's code without further
permission, including changing it for one's own use and distributing
it again.
\item Open Standards. One can find visible community mechanisms for
protocols and communicating information. The mechanisms for creating
and maintaining these standards cover a wide spectrum of human
organisations, including various degrees of consent.
\item Open Data. One can obtain all data in the public domain when
wanted and reuse it for whatever purpose.
\end{itemize}

The Blue Obelisk was first described in the CDK News \cite{CDKNewsBO} and
later as a formal paper by Guha et al.\cite{Guha2006} in
2006. This
contribution looks back on the past 5 years and surveys progress and
remaining challenges in the areas of Open Source, Open Data and Open
Standards in chemistry.

@@

Some project have also been adopted by other fields, including biology
and crystallography. A good example
here is Jmol, which has, in the past five years, gone through two major revisions
as part of an accelerated development focus. Starting with Jmol 10.2, released in
April of 2006, Jmol 11.0 added over 100 new features and was released in March of
2007. Three years later, Jmol 12.0 followed with over 700 new features.
Based on the outstanding work of Michael (Miguel) Howard and others, this development
path took Jmol from a ``Rasmol/Chime replacement'' to a fully fledged molecular
visualisation package, including full support for crystallography~\cite{Hanson2010},
efficient delivery of surface data using fast solvent-excluded surface
generation and the highly compressed JVXL format [http://chemapps.stolaf.edu/jmol/docs/misc/JVXL-format.pdf],
 display of molecular orbitals from standard basis set/coefficient data,
the inclusion of dynamic minimisation using the UFF force field, and
a full implementation of Daylight SMILES and SMART, with extensions to
conformational and biomolecular substructure searching in the form of Jmol
BioSMARTS.[http://jmol.svn.sourceforge.net/viewvc/jmol/trunk/Jmol/src/org/jmol/smiles/package.html]
Behind the scenes, Jmol has been essentially completely rewritten. A full
mathematical expression evaluator was added, and the scripting language
was folded into a versatile JavaScript-like command language, including a
complete set of program flow commands, several new data types, and
user-defined variables and functions. An exciting non-chemical application
of Jmol has been its incorporation into the open source Sage
Mathematics Project [http://www.sagemath.org/], where Jmol is being used
to deliver three-dimensional depiction of mathematical functions.

Two new Open Source cheminformatics toolkits have appeared since the
original paper. In 2006 Rational Discovery, a cheminformatics service
company (since closed down), released RDKit \cite{WebRDKit} under the
BSD License. RDKit continues to be actively developed and includes
code donated by Novartis.

More recently, GGA Software Services (a contract programming company)
released the Indigo toolkit \cite{WebIndigo} and associated software
in 2009 under the GPL. Indigo is written from scratch in C++ and has
high-level wrappers for using it in C, Java, Python, or .NET
environment. Like other toolkits, Indigo provides support for
tetrahedral and cis-trans stereochemistry, 2D coordinate generation,
exact/substructure/SMARTS matching, fingerprint generation (all these
both for molecules and reactions), and canonical SMILES computation.
It also provides some less common functionality, like matching
tautomers and resonance substructures, enumeration of subgraphs,
finding maximum common substructure of $N$ input structures, and
enumerating reaction products. Indigo also has its rendering plugin,
which is used in a lot of commercial and open-source projects as the
main 2D chemical rendering engine.

While desktop software has composed the majority of scientific tools since the computer was introduced, the internet continues to change how applications and content are distributed and presented. As such, there is a growing need to present chemical and scientific information in a web-based medium. Recently, a new version of the HTML specification, HTML5 \cite{html5}, defines a well-developed framework for creating native web applications in JavaScript. This native framework provides a significant benefit to scientists as the web is an open and free medium to distribute scientific knowledge, ideas and education. It also poses significant new problems, as the cost associated with learning and developing tools for new technologies is not trivial. As such, it is an important goal for the Blue Obelisk to provide such tools and encourage the community to utilise these new technologies to improve products and education. A specific example of this initiative is the ChemDoodle Web Components library \cite{ChemDoodleWeb}, produced by iChemLabs. The ChemDoodle Web Components library is open source under the GPL v3 license with a liberal HTML exception. ChemDoodle Web Components allow the scientist to present publication quality 2D and 3D graphics and animations for chemical structures, reactions and spectra. Beyond graphics, this tool provides a framework for user interaction to create dynamic applications through web browsers, desktop platforms and mobile devices such as the iPhone, iPad and Android devices. It is now easier than ever for scientists to create web-based scientific content, without having to master complicated technologies such as Java applets. iChemLabs is dedicated to funding, developing and supporting the library to ensure that the next generation of scientific applications is easily achievable by academia, government and industry. This goal helps to make sure that the cost of education decreases while using the web to further spread science.

@@

Bioclipse~\cite{Spjuth:2007fk} (v2.4 released in Aug 2010) and Avogadro
\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such software, based
on the CDK and Open Babel, respectively. Bioclipse is an award-winning
molecular workbench for life sciences with a dynamic plugin architecture, and
which wraps cheminformatics functionality behind user-friendly interfaces and
graphical editors. All functionality in Bioclipse is completely scriptable
using an integrated high-level language called Bioclipse Scripting Language
(BSL), which can be used for scripted research or to automate complex
tasks~\cite{Bioclipse2}. Bioclipse in turn acts as base for other software
such as Brunn~\cite{Brunn}, which is a laboratory information system for
microplate based high-throughput screening. In this manner open source projects
empower other open source projects.

Avogadro is a 3D molecular editor and viewer aimed at preparing
and analysing computational chemistry calculations. An interesting
aspect of both Avogadro and Bioclipse is that they share some developers
with the underlying toolkits and this has driven the development of new
features in the CDK and Open Babel.

CDK-Taverna \cite{Kuhn:2010p4001} as an open-source cheminformatics workflow solution.
Commonly used in this context are workflow engines for cheminformatics, where numerous recurring tasks can be automated, including tasks for chemical data filtering, transformation, curation and migration workflows, chemical documentation and information retrieval related workflows (structures, reactions, pharmacophores, object relational data etc.) or data analysis workflows (statistics and clustering/ machine learning for QSAR, diversity analysis etc. 
CDK-Taverna has matured to become a freely available and increasingly powerful tool for the biosciences. It was recently ported to 
Taverna version 2.x.
The combination of CDK-Taverna workers with the multitude of workers already published on myexperiment.org by the large and active user community of Taverna enables scientists to quickly build workflows to process a diverse set of data as typically found in today's systems biology scenarios.

@@

Open Source provides a unique opportunity for commercial organisations to work with the community. As opposed to traditional business models which focus on maximising revenue through monetisation of all aspects of software, working with the community allows a business to maximise profits through greater community adoption and popularity. This is an ideal solution for some commercial organisations because only they can afford the significant investment in funding and time that is required to develop a successful open source project. Regardless, there are always challenges to growing a project, and there are specific issues that complicate making a business out of open source software in the scientific industry. A goal of the Blue Obelisk is to reduce these challenges and bring commercial organisations and the community together to benefit science.

Open source products attract interest because they are easily modifiable, and they can usually be used freely under a standard open source license. With proper advertising, a company can grow a user base more quickly than with a proprietary product, which requires significantly more marketing resources. Additionally, an open source product may integrate with other products that are governed by compatible open source licensing, allowing for faster development. Adopters benefit from a quickly established community of users, that not only know how to use that software, but may also have experience developing it. These benefits allow a company to create a successful product with significantly reduced cost and effort.

Successful open projects require the dedication of a number of individuals to ensure that the product is continually developed, supported and funded. In many non-commercial cases, an author will produce an open project as a hobby, and stop work on the project before the development reaches a certain quality or community interest reaches a critical mass. This is because the hobbyist eventually needs to focus on income or he/she has a change of priorities. On the other hand, commercial organisations have the means to devote resources to a project to ensure that it achieves an appropriate level of quality and adoption. In return, that commercial organisation gains recognition from that product and is the first choice for commercial support, consulting and custom development. This is the primary goal of commercial open source, as selling licenses for an open product is very unlikely to be feasible.

In the scientific industry, it is incredibly difficult to start successful open projects. This is because the expertise needed to contribute to such a project requires a higher degree of education and the population of end users is much smaller than other industries, such as entertainment. Because of this, there is an increased barrier to starting open projects in the scientific industry. The Blue Obelisk attempts to solve this issue, by uniting those interested in open source and by helping to increase awareness for new open projects. Commercial organisations can take advantage of this goal, as open products will provide a faster return on investment, and the community benefits from open software and new job opportunities.

For instance, iChemLabs distributes several scientific software solutions under various open and proprietary licensing. The license is chosen to best ensure the success of the project. iChemLabs released the ChemDoodle Web Components library under the GPL v3 license with a liberal HTML exception. iChemLabs is dedicated to developing, funding and supporting the library so that the scientific community can immediately utilise the maturing HTML5 technologies and quickly make web and mobile interfaces for scientific content. Other open source developers can benefit from this library freely and expect continued development, while commercial organisations that use the library under an appropriate license (GPL for open projects or purchasing a proprietary license for proprietary projects) will benefit from an already developed market. Of course, no one will need to waste time rewriting functionality that is already provided with quality. Several open projects have already incorporated the ChemDoodle Web Components, such as iBabel by Chris Swain\cite{iBabel}, ChemSpotlight by Geoffrey Hutchison\cite{chemspotlight} and the RSC ChemSpider\cite{chemspider_chemdoodle}.

An example of such an interaction was the donation by eMolecules, Inc.
to Open Babel of code for the canonicalisation of molecules and
fragments (Nov 2006). eMolecules is an online vendor of chemicals that
uses Open Babel under-the-hood to manage its compound collections.
Another more recent example occurred in July 2010 when Silicos, a
Belgian company that provides services
in the area of cheminformatics, released several command line
applications based on Open Babel as well as donating code to the
project. For example, the Pharao
tool released by them is a comprehensive solution for pharmacophore
searching that provides extensive support for a variety of
pharmacophore searches. Other tools released by them include Sieve for
filtering by molecular property and XXX.

@@

One of the effects of the Blue Obelisk has been to bring developers
together from different Open Source chemistry projects so that they
look for opportunities to collaborate rather than compete, and to
leverage work done by other projects to avoid duplication of effort.
As an example of this, when in March 2008 the Jmol development team
were looking to add support for energy minimisation, rather than
implement a forcefield from scratch they ported the UFF forcefield
implementation from Open Babel to Jmol. This code has allowed Jmol to
support 2D to 3D conversion of structures (through energy
minimisation). Similarly, efficient Jmol code for atom-atom rebonding
has been ported to the CDK.

@@

In the area of cheminformatics toolkits, two of the existing toolkits
Open Babel and RDKit are planning to work together on a common
underlying framework called MolCore. This project is still in the
planning stage, but if it is a success it will mean that the two
libraries will be interoperable (while retaining their existing focus)
but also that the cost of maintaining the code will be shared among
more developers, freeing time for the development of new features.

@@

Performance. There is a famous quote by Knuth that "Premature
optimisation is root of all evil (or something)". Given that compute
time is cheap, ....

\section*{Chemical Structure Registration Systems, Databases}
Databases? (XXX CS: A couple of database-related paragraphs could go here XXX)

XXX the paragraph below on OrChem needs some work
Registration, indexing and searching of chemical structures in relational databases is one of the core areas of cheminformatics. 
A number of structure registration systems have been published in the last five years, exploiting the fact that 
free cheminformatics toolkits such as OpenBabel and the CDK were available. 
OrChem, for example, is an extension for the Oracle 11G database that adds registration and indexing of chemical structures to support fast substructure and similarity searching. The cheminformatics functionality is provided by the Chemistry Development Kit. OrChem provides similarity searching with response times in the order of seconds for databases with millions of compounds, depending on a given similarity cut-off. For substructure searching, it can make use of multiple processor cores on today's powerful database servers to provide fast response times in equally large data sets.
OrChem is free software and can be redistributed and/or modified under the terms of the GNU Lesser General Public License as published by the Free Software Foundation. All software is available via http://orchem.sourceforge.net.

@@

Also, Indigo is going to provide its own plugin with an interface to
the InChI implementation in May 2011. This will enable developers to
calculate InChI and InChIKey-s within Python and .NET projects (not to
mention Java), eliminating the trouble of compiling and using binary
modules directly.

@@

Recently proposed CurlySMILES\cite{CurlySMILES} is an extension of the
SMILES notation, which allows to define crystals structures, polymers,
electron delocalisation charges, molecule interactions, and many other
features absent in the initial SMILES specification. There has been
discussions about including parts of the CurlySMILES notation into
OpenSMILES, especially polymers.

\subsubsection*{CML}
?

    \subsubsection*{IUPAC Nomenclature Conversion}
The oldest form of communication used to describe chemicals, predating even the knowledge of the atomic structure of compounds, is the chemical name. Even now chemical names are abundant in the scientific literature and encode valuable structural information. The IUPAC have through successive books of recommendations~\cite{iupac_nomenclature_1979, iupac_guide_1993} aimed to codify and to an extent standardise naming practices. The latest provisional recommendations are freely available through the IUPAC's website~\cite{iupac_draft_2004}.

The Blue Obelisk project OPSIN (Open Parser for Systematic IUPAC Nomenclature)~\cite{lowe_chemical_2011} aims to make this abundance of chemical names machine readable by translating them to SMILES, CML or InChI. The program is based around the use of a regular grammar to guide tokenisation and parsing of chemical names, followed by step-wise application of nomenclature rules. OPSIN is able to offer fast and precise conversions for the majority of names using IUPAC organic nomenclature. OPSIN is available as a web service, Java library and standalone application for maximum interoperability.

\subsubsection*{QSAR-ML}
The field of QSAR has long been hampered by the lack of open standards, which makes it difficult to share and reproduce descriptor calculations and analyses. QSAR-ML was recently proposed as an open standard for exchanging QSAR datasets~\cite{Spjuth:2010uq}. A dataset in QSAR-ML includes the chemical structures (preferable described in CML) with InChI to protect integrity, chemical descriptors by linking to the Blue Obelisk Descriptor Ontology~\cite{bodo}, response values, units, and versioned descriptor implementations to allow for integrating several descriptor software in the same calculation. Hence, a dataset described in QSAR-ML is completely reproducible. To allow for easy setup of QSAR-ML compliant datasets, a plugin for Bioclipse was created with graphical interfaces that can be used to set up QSAR datasets and perform calculations. Descriptor implementations were initially available from CDK and JOELib, as well as via remote web services such as XMPP~\cite{Wagener:2009uq}.

\subsubsection*{Others...?}

@@

A core requirement for chemical structure databases and chemical
registration systems in general is the notion of structure
standardisation.  That is,  for a given input structure, multiple
representations should be converted to one canonical form. 
Structure canonicalisation routines partially address this aspect,
converting multiple alternative topologies to a single canonical
form. However, the problem of standardisation is broader than just
topological canonicalisation. Features that must be considered include
\begin{itemize}
\item topological canonicalisation
\item handling of charges
\item tautomer enumeration and canonicalisation
\item normalisation of functional groups
\end{itemize}
Currently, most of the individual components of a `standardisation
pipeline' can be implemented using BO tools. The larger problem is
that there is no agreed upon list of steps for a standardisation
process. While some specifications have been published (e.g., PubChem)
and some standardisation services and tools are available (PubChem
provides an online service to standardise molecules and the NCGC
provides a stand alone tool) each group has their own set of rules. A
common reference specification for standardisation would be of immense
value in interoperability between structure repositories as well as
between toolkits (though the latter is still confounded by differences
in lower level cheminformatic features such as aromaticity models).

@@

A considerable stumbling block in advocating the release of scientific
data as Open Data has been how exactly to define Open. A major step
forward was the launch in 2010 of the Panton Principles for Open Data
in Science \cite{WebPanton}. This formalises the idea that Open Data maximises the
possibility of reuse and repurposing, the fundamental basis
of how science works. These principles recommend that published data
be licensed explicitly, and preferably under CC0 (Creative Commons `No
Rights Reserved', also known as CCZero) \cite{WebCC0}. This license allows others to use the 
data for any purpose whatsoever without any barriers. Other licenses
compatible with the Panton Principles include the
Open Data Commons Public Domain Dedication and Licence (PDDL), the
Open Data Commons Attribution License, and the
Open Data Commons Open Database License (ODbL) (see
http://www.opendefinition.org/licenses/\#Data).

Despite this positive news, little chemical data has become
available from the traditional chemical fields of organic,
inorganic, solid state chemistry. Table~2 lists a few notable
exceptions: ChemPedia (a now discontinued crowd-sourcing project),
CrystalEye (http://wwmm.ch.cam.ac.uk/crystaleye/),
and the Open Notebook Science Solubility
data~\cite{ONS2010}. There is also data available using licenses
not compatible with the Panton Principles, but where the user
is allows to modify and redistribute the data. A new data
set in this category is the data from the ChEMBL database, 
which is available under the Creative Commons Share-Alike
Attribution license~\cite{Overington2009}.

Importantly, publishing data as CC0 is become easier now that
website are becoming available to simplify publishing data. Two
projects that can be mentioned in this context are FigShare
(http://figshare.com/), where the data behind unpublished figures
can be hosted, and Dryad (http://datadryad.org/) where data
behind publications can be hosted. Initiatives like this make
it possible to host small amounts of data, and those combined
are expected to become soon a substantial knowledge base.

@@

\section*{Other areas of activity}
Web services? XMPP services? (NIH, CDK and ChemSpider)

\subsection*{Getting the message out}

One recent development
that attempts to address this issue was the establishment in Apr 2010 of a
question-and-answer (Q\&A)
website related to Blue Obelisk projects and themes at
http://blueobelisk.shapado.com (see Figure~3). This is a website in the
style of Stack Overflow that encourages high quality answers (and
questions) through the use of a voting system. In the year since it
was established, over
200 users have registered, many of whom had no previous involvement
with the Blue Obelisk, showing that the Q\&A website is a nice complement
to earlier existing channels of communication.

@@

We hope to see this involvement between the Blue Obelisk and the wider
community grow in the future. To this end, we encourage the reader to
visit http://blueobelisk.org, send a message to our mailing list,
investigate related projects or read our blogs.

@@

  \subsection*{Figure 3 - Screenshot of the Blue Obelisk eXchange Question
    and Answer website.}

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Blue Obelisk Open Source software projects}
    (Description if necessary XXXXXXXXXXXXXXX. Add citations to project names.)
    \par \mbox{}
    \par
    \mbox{
      \begin{tabular}{|c|c|c|}
        %% \hline \multicolumn{3}{|c|}{My Table}\\ \hline
        \hline Name & Website & Description? or Lead Developer? \\ \hline
        \multicolumn{3}{|c|}{Cheminformatics toolkits} \\ \hline
        Chemistry Development Kit (CDK) \cite{Steinbeck2003, Steinbeck2006} & http://cdk.sf.net & XXXX  \\ \hline
        Cinfony & http://cinfony.googlecode.com & Noel O'Boyle, Python interface to toolkits \\ \hline
        Indigo & http://ggasoftware.com/opensource/indigo & GGA Software \\ \hline
        Open Babel & http://openbabel.org & Geoffrey Hutchison et al \\ \hline
        RDKit & http://rdkit.org & Greg Landrum \\ \hline
        ChemDoodle Web Components & http://web.chemdoodle.com & iChemLabs \\ \hline
        \multicolumn{3}{|c|}{Integration} \\ \hline
        CDK-Taverna \cite{Kuhn:2010p4001} & http://cdk-taverna.blah.XXX & Christoph Steinbeck, Workflow \\ \hline
        A3 & ..  & .  \\ \hline
        \multicolumn{3}{|c|}{Interconversion} \\ \hline
        OSRA & http://osra.sf.net & Igor Filippov, Image to structure \\ \hline
        OPSIN & http://opsin.ch.cam.ac.uk & Daniel Lowe, Name to structure \\ \hline
        A3 & ..  & .  \\ \hline
        \multicolumn{3}{|c|}{Structure Databases} \\ \hline
        OrChem \cite{RijnbeekS10} & http://orchem.sourceforge.net & Christoph Steinbeck, Oracle-based Chemical Database Engine \\ \hline
        MyChem  & http://XXX.sourceforge.net & Jerome Pansanel, MySQL-based Chemical Database Engine \\ \hline
        PGChem  & http://XXX.sourceforge.net & Ernst-Georg Schmid, PostgreSQL-based Chemical Database Engine \\ \hline
        Bingo  & http://XXX.XXX.XXX & Dmitry Pavlov, Oracle-based Chemical Database Engine \\ \hline

>>>>

\author{Charles A Darwin\correspondingauthor$^{1,2}$%
       \email{Charles A Darwin\correspondingauthor - charles@londonzoo.co.uk}%
      \and 
         Egon L Willighagen$^3$%
         \email{Egon L Willighagen - egon.willighagen@ki.se}
     \and 
         Rajarshi Guha$^4$%
         \email{Rajarshi Guha - guhar@mail.nih.gov}
     \and 
         Noel M O'Boyle$^5$%
         \email{Noel M O'Boyle - n.oboyle@ucc.ie}
      \and
         Jane E Doe\correspondingauthor$^2$%
         \email{Jane E Doe\correspondingauthor - jane.e.doe@cambridge.co.uk}
      }

@@

\address{%
    \iid(1)Life Sciences Department, Kings College London, Cornwall House,%
        Waterloo Road, London, UK\\
    \iid(2)Department of Zoology, Cambridge, Waterloo Road, London, UK\\
    \iid(3)Division of Molecular Toxicology, Institute of Environmental Medicine, %
        Nobels vaeg 13, Karolinska Institutet, 171 77 Stockholm, Sweden
    \iid(5)Analytical and Biological Chemistry Research Facility, Cavanagh Pharmacy Building, University College Cork, College Road, Cork, Co. Cork, Ireland
}%

@@

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
        \paragraph*{Background:} Text for this section of the abstract.

        \paragraph*{Results:} Text for this section of the abstract \ldots

        \paragraph*{Conclusions:} Text for this section of the abstract \ldots
\end{abstract}

@@

%%%%%%%%%%%%%%%%
%% Background %%
%%
\section*{Background}
The Blue Obelisk group was established in 2005 at the
229\textsuperscript{th} National Meeting of the American Chemistry
Society as a response to the lack of open data, open standards and
open source (ODOSOS) in chemistry. While other scientific disciplines
such as physics, biology and astronomy (to name a few) were embracing
new ways of doing science and reaping the benefits of community
efforts, there was little if any innovation in the field of chemistry
and scientific progress was actively hampered by the lack of access to
data and tools.

The formation of the Blue Obelisk group is somewhat unusual in that it
is not a funded network, nor does it follow the industry consortium
model. Rather it is a grassroots organisation, catalysed by an initial
core of interested scientists, but with membership open to all who
share one or more of the goals of the group:
\begin{itemize}
\item Open Source. One can use other people's code without further
permission, including changing it for one's own use and distributing
it again.
\item Open Standards. One can find visible community mechanisms for
protocols and communicating information. The mechanisms for creating
and maintaining these standards cover a wide spectrum of human
organizations, including various degrees of consent.
\item Open Data. One can obtain all data in the public domain when
wanted and reuse it for whatever purpose.
\end{itemize}

Note that while some may advocate also for Open Access to
publications, the goals focus more on the availability of code (to
reproduce results), standards (to exchange data), and the scientific
data itself. All three of these goals stem from the basics of the
scientific method for data sharing and reproducibilty.

The Blue Obelisk was first described in Guha et al.\cite{Guha2006} in
2006. This
contribution looks back on the past 5 years and surveys progress and
remaining challenges in the areas of Open Source, Open Data and Open
Standards in chemistry.

@@

Two new Open Source cheminformatics toolkits have appeared since the
original paper. In 2006 Rational Discovery, a cheminformatics service
company (since closed down), released RDKit \cite{WebRDKit} under the
BSD License. RDKit continues to be actively developed and includes
code donated by Novartis. More recently, GGA Software Services (a
contract programming company) released the Indigo toolkit
\cite{WebIndigo} and associated software in 2009 under the GPL. [More
text on Indigo here]

One important aspect of the above
toolkits is that they originated for the support of desktop
and server-side applications. More recently, a number of workers have developed light-weight
toolkits that are targeted towards web applications. One example of
this form is the ChemDoodle library from iChemLabs. The focus of these
toolkits is to support client-side cheminformatics functionality,
allowing for rich, chemically-aware browser-based applications.

@@

Bioclipse \cite{Bioclipse2} (v1.0 released in Aug 2006) and Avogadro
\cite{WebAvogadro} (v1.0 in Oct 2009) are two examples of such
software, based on the CDK and Open Babel, respectively. Bioclipse is
an award-winning `molecular workbench' for life sciences that ...,
while Avogadro is a 3D molecular editor and viewer aimed at preparing
and analysing computational chemistry calculations. An interesting
aspect of both of these projects is that they share some developers
with the underlying toolkit and this has driven the development of new
features in the CDK and Open Babel.

@@

Although the software developed by the Blue Obelisk is Open Source on
principle, this does not rule out the establishment of a business
based on open source chemistry software. Rather than supporting itself
through sales, such a business generally relies on providing support,
customisation and services around the software. The interaction
between an open source project and a company is important, and works
best where there is some form of `giving back' to the project (such as
developer time, code contributions, or financial assistance with server costs).

An example of such an interaction was the donation by eMolecules, Inc.
to Open Babel of code for the canonicalisation of molecules and
fragments (Nov 2006). eMolecules is an online vendor of chemicals that
uses Open Babel under-the-hood to manage its compound collections.

Another more recent example occurred in July 2010 when Silicos, a
Belgian company that provides services
in the area of cheminformatics, released several command line
applications based on Open Babel as well as donating code to the
project. For example, the Pharao
tool released by them is a comprehensive solution for pharmacophore
searching that provides extensive support for a variety of
pharmacophore searches. Other tools released by Silicos include Sieve for
filtering by molecular property, Stripper for removing core scaffold
structures from a molecule set, and Piramid for molecular alignment
using shape determined by the Gaussian volumes as a descriptor.

@@

One of the effects of the Blue Obelisk has been to bring developers
together from different Open Source chemistry projects so that they
look for opportunities to collaborate rather than compete, and to
leverage work done by other projects to avoid duplication of effort.
As an example of this, when in March 2008 the Jmol development team
were looking to add support for energy minimisation, rather than
implement a forcefield from scratch they ported the UFF forcefield
implementation from Open Babel to Jmol. This code has allowed Jmol to
support 2D to 3D conversion of structures (through energy
minimisation).

@@

In the area of cheminformatics tookits, two of the existing toolkits
Open Babel and RDKit are planning to work together on a common
underlying framework called MolCore.\cite{WebMolCore} This project is still in the
planning stage, but if it is a success it will mean that the the two
libraries will be interoperable (while retaining their existing focus)
but also that the cost of maintaining the code will be shared among
more developers, freeing time for the development of new features.

@@

Performance. There is a famous quote by Knuth that "Premature
optimization is root of all evil (or something)". Given that compute
time is cheap, ....

Databases?

@@

@@

CML?

@@

A core requirement for chemical structure databases and chemical
registration systems in general is the notion of structure
standardization.  That is,  for a given input structure, multiple
representations should be converted to one canonical form. 
Structure canonicalization routines partially address this aspect,
converting multiple alternative topologies to a single canonical
form. However, the problem of standardization is broader than just
topological canonicalization. Features that must be considered include
\begin{itemize}
\item topological canonicalization
\item handling of charges
\item tautomer enumeration and canonicalization
\item normalization of functional groups
\end{itemize}
Currently, most of the individual components of a `standardization
pipeline' can be implemented using BO tools. The larger problem is
that there is no agreed upon list of steps for a standardization
process. While some specifications have been published (e.g., Pubchem)
and some standardization services and tools are available (Pubchem
provides an online service to standardise molecules and the NCGC
provides a stand alone tool) each group has their own set of rules. A
common reference specification for standardization would be of immense
value in interoperability between structure repositories as well as
between toolkits (though the latter is still confounded by differences
in lower level cheminformatic features such as aromaticity models).

@@

A considerable stumbling block in advocating the release of scientific
data as Open Data has been how exactly to define Open. A major step
forward was the launch in 2010 of the Panton Principles for Open Data
in Science \cite{WebPanton}. This formalises the idea that Open Data maximises the
possibility of reuse and repurposing, the fundamental basis
of how science works. These principles recommend that published data
be licensed explicitly, and preferably under CC0 (Creative Commons `No
Rights Reserved') \cite{WebCC0}. This license allows others to use the data for any
purpose whatsoever without any barriers.

@@

\section*{Other areas of activity}
Web services? (NIH, CDK and Chemspider)

@@

One recent development
that attempts to address this issue was the establishment in Apr 2010 of a
question-and-answer
website related to Blue Obelisk projects and themes at
http://blueobelisk.shapado.com. This is a website in the
style of Stack Overflow that encourages high quality answers (and
questions) through the use of a voting system. In the year since it
was established, over
200 users have registered, many of whom had no previous involvement
with the Blue Obelisk.

We hope to see this involvement between the Blue Obelisk and the wider
community grow in future. To this end, we encourage the reader to
visit http://blueobelisk.org, send a message to our mailing list,
investigate related projects or read our blogs.

@@

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
  \subsection*{Table 1 - Blue Obelisk Open Source software projects}
    (Description if necessary XXXXXXXXXXXXXXX. Add citations to project names.)
    \par \mbox{}
    \par
    \mbox{
      \begin{tabular}{|c|c|c|}
        %% \hline \multicolumn{3}{|c|}{My Table}\\ \hline
        \hline Name & Website & Description? or Lead Developer? \\ \hline
        \multicolumn{3}{|c|}{Cheminformatics toolkits} \\ \hline
        Chemistry Development Kit & http://cdk.sf.net & XXXX  \\ \hline
        Cinfony & http://cinfony.googlecode.com & Noel O'Boyle, Python interface to toolkits \\ \hline
        Indigo & http://ggasoftware.com/opensource/indigo & GGA Software \\ \hline
        Open Babel & http://openbabel.org & Geoffrey Hutchison et al \\ \hline
        RDKit & http://rdkit.org & Greg Landrum \\ \hline
        \multicolumn{3}{|c|}{Interconversion} \\ \hline
        OSRA & http://osra.sf.net & Igor Filippov, Image to structure \\ \hline
        A3 & ..  & .  \\ \hline
      \end{tabular}
      }
  \subsection*{Table 2 - Sample table title}
    Large tables are attached as separate files but should
    still be described here.
